Rao-Blackwellized smoother for Conditionally Linear and Gaussian models

### 2016-11-12 TexArxiv/clgm.tex
Numerical results updated

p^{\mathsf{Lbscg}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.

@@

y^k_{i|i-1}(a_i) = c_{a_i} + B_{a_i}\mu^k_{i-1}\;\mbox{and}\; \Sigma^k_{i|i-1}(a_i) = B_{a_i}P^k_{i-1}B'_{a_i} + \barG_{a_i}\eqsp.

@@

\paragraph{Additional Rao-Blackwellisation step:}

@@

\section{Numerical experiments}
\label{sec:numerical:experiments}
This section highlights the performance of the Rao Blackwellised Two-Filter approximation (RBTF) of the marginal smoothing distributions in comparison to the other Rao Blackwellised smoothers introduced in Section~\ref{sec:RaoBlackwell}.

\subsection{Simulated data}
We consider the two-factor model \eqref{eq:model:state} and \eqref{eq:model:obs} with $J=2$. The performance of the algorithms are analyzed in two different settings with $n=1500$ observations. In both settings the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}. The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.  It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$). The initial state variance is high to emphasize the fact that there is no prior precise knowledge on the state value.

\paragraph{Experiment 1:}
\[
\pi(1)=\pi(2) = 1/2\\;\;d_1 = \begin{pmatrix}0 \\ 0.5 \end{pmatrix}\;\; d_2 = \begin{pmatrix}0 \\ 0\end{pmatrix}\\;\;c_1= \begin{pmatrix}0 & 0& 0 & 0 \end{pmatrix}'\;\;c_2 =\begin{pmatrix}0 & 0 & 0 & 0 \end{pmatrix}'\eqsp,
\]
\[
Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T = \begin{pmatrix}1 & -0.02\\ 0 & 0.96\end{pmatrix}\;\;\barH_1 = \begin{pmatrix}0.25 & 0.01\\ 0.01 & 0.01\end{pmatrix}\;\;\barH_2 = \begin{pmatrix}0.3 & 0\\ 0 & 0.3\end{pmatrix}\eqsp,
\]
\[
B = \begin{pmatrix}1 & 1 & 1 & 1\\ 0 & 0.5 & 0.5 & 0.1 \end{pmatrix}'\;\; \barG = \mathrm{diag}\left(0.1,0.1,0.1,0.1\right)\;\; \mu_1 = \begin{pmatrix}1 \\ 0\end{pmatrix}\;\; \Sigma_1 = \begin{pmatrix} 5 & 0 \\ 0  & 5\end{pmatrix}\eqsp.
\]
The approximations of the smoothing distributions are computed with $N=10$. RBTF and the algorithm of \cite{briers:doucet:maskell:2010} are compared to the approximations provided by RBTF with $1000$ particles.
%Figure~\ref{fig:exp1:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.

@@

%\begin{figure}
%\centering
%\includegraphics[scale=.5]{Artificial_Compare_Exp1_1_500.eps}
%\includegraphics[scale=.5]{Artificial_Compare_Exp2_1_500.eps}
%\caption{estimated probabilities by RBTF in experiment 1 [top] and experiment 2 [bottom] with the two artificial laws: yellow (the first one) and red (the third one).}
%\label{fig:exp:artificial}
%\end{figure}

Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the labels given by the algorithms using:
\[
\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
\]
In both cases, RBTF behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).

>>>>

p^{\mathsf{Lbscg}}_{\tilde{N}}(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.

@@

y^k_{i|i-1}(a_i) = c_{a_i} + B_{a_i}(d_{a_i}+T_{a_i}\mu^k_{i-1})\;\mbox{and}\; \Sigma^k_{i|i-1}(a_i) = B_{a_i}(T_{a_i}P^k_{i-1}T'_{a_i}+\barH_{a_i})B'_{a_i} + \barG_{a_i}\eqsp.

@@

@@

\section{Numerical experiments}
\label{sec:numerical:experiments}
\subsection{Simulated data}
This section highlights the performance of the additional Rao Blackwellisation steps for the Two-Filter and the FFBS approximations of the marginal smoothing distributions in the case where $J=2$. In each setting the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}. First, the FFBS algorithm is applied to a simple one-dimensional model with:
\[
\pi(1)=\pi(2) = 1/2\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= c_2 = 0 \eqsp,
\]
\[
Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T_1 = T_2 = 1\;\;\barH_1 = \barH_2= 0.05\eqsp,
\]
\[
B_1 = B_2 =1\;\;\barG_1 = 0.3\;\;\barG_2 = 0.1\eqsp.
\]
The two FFBS algorithms are used with $N = \tilde{N} = 100$.
Figure~\ref{fig:ffbs} displays the RMSE values as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm. Then, the performance of the two-filter based algorithms are compared in two different settings with $n=1500$ observations.
\paragraph{Experiment 1:}
\[
\pi(1)=\pi(2) = 1/2\\;\;d_1 = \begin{pmatrix}0 \\ 0.5 \end{pmatrix}\;\; d_2 = \begin{pmatrix}0 \\ 0\end{pmatrix}\\;\;c_1= \begin{pmatrix}0 & 0& 0 & 0 \end{pmatrix}'\;\;c_2 =\begin{pmatrix}0 & 0 & 0 & 0 \end{pmatrix}'\eqsp,
\]
\[
Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T = \begin{pmatrix}1 & -0.02\\ 0 & 0.96\end{pmatrix}\;\;\barH_1 = \begin{pmatrix}0.25 & 0.01\\ 0.01 & 0.01\end{pmatrix}\;\;\barH_2 = \begin{pmatrix}0.3 & 0\\ 0 & 0.3\end{pmatrix}\eqsp,
\]
\[
B = \begin{pmatrix}1 & 1 & 1 & 1\\ 0 & 0.5 & 0.5 & 0.1 \end{pmatrix}'\;\; \barG = \mathrm{diag}\left(0.1,0.1,0.1,0.1\right)\;\; \mu_1 = \begin{pmatrix}1 \\ 0\end{pmatrix}\;\; \Sigma_1 = \begin{pmatrix} 5 & 0 \\ 0  & 5\end{pmatrix}\eqsp.
\]
The approximations of the smoothing distributions are computed with $N=10$. RBTF and the algorithm of \cite{briers:doucet:maskell:2010} are compared to the approximations provided by RBTF with $1000$ particles.
%Figure~\ref{fig:exp1:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.

@@

%\begin{figure}
%\centering
%\includegraphics[scale=.5]{Artificial_Compare_Exp1_1_500.eps}
%\includegraphics[scale=.5]{Artificial_Compare_Exp2_1_500.eps}
%\caption{estimated probabilities by RBTF in experiment 1 [top] and experiment 2 [bottom] with the two artificial laws: yellow (the first one) and red (the third one).}
%\label{fig:exp:artificial}
%\end{figure}
The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.  It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$). Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the labels given by the algorithms using:
\[
\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
\]
In both cases, RBTF behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
